Parallel Quadtree Construction on Collections of Objects

### 2017-04-25 paper.tex
Updated paper to take into account the clarity concerns addressed by Reviewer 3. Formatting's a little messed up now.

\paragraph{Parallel} Many recent works on fast quadtree construction using the GPU are limited to point sites \cite{bedorf2012sparse,karras2012maximizing,zhou2011data}.
%Kim and Liu's work~\shortcite{kim2014exact} is similar, computing the quadtree on the barycenters of triangles, giving an approximation of an object-resolving quadtree.
 Most quadtree approaches that support surfaces \cite{baert2013out,crassin2009gigavoxels,laine2011efficient,lefebvre2007compressed} are designed for efficient rendering, and actual construction of the quadtree is implemented on the CPU. Two works \cite{bastos2008gpu,park2010cuda} implement Adaptive Distance Fields in parallel on quadtrees but building the quadtree itself is done sequentially.  Yin \etal~\shortcite{yin2011fast} compute the octree entirely on the GPU using a bottom-up approach by initially subdividing into a complete quadtree, resulting in memory usage that is no better than using a uniform grid.
The most similar work to what we do here is Kim and Liu's method~\shortcite{kim2014exact}, which computes the quadtree on the barycenters of triangles, giving an approximation of our quadtree, but without fully resolving between objects.
We have found no GPU quadtree construction method that is fully adaptive and can resolve between objects.

@@

\subsection{Build initial quadtree}
\label{sec:build-initial-quadtree}
Our first step is to build a quadtree on the given set of vertices. We use the Karras algorithm \cite{karras2012maximizing} which begins by placing the given vertices on a Z-Order curve by computing each vertex's cooresponding Morton code in parallel. Next, Karras sorts the converted points by using a parallel radix sorter, which has a linear execution time. Our implementation uses the efficient four-way parallel radix sorter described by Ha \etal \shortcite{ha2009fast}. Once the morton codes are sorted, the Z-Order curve can be exploited to construct a binary radix tree in a parallel bottom up manner by identifying longest common Morton code prefixes between neighboring points. This resultant binary radix tree can be analyzed in parallel to identify the size and structure of the required vertex quadtree. The strength of this approach lies in the fact that overall performance scales linearly with the number of cores, regardless of the distribution of points. That is, even if a large number of vertices are clustered in a small area, requiring deep quadtree subdivision, only a constant number of parallel calls need be made.

@@

\subsection{Pruning the quadtree}
\label{sec:pruning}
During Karras' initial binary radix tree (BRT) construction, we can prune the BRT to simplify the resultant quadtree. This in turn simplifies the work complexity of conflict cell detection and reduces our overall memory footprint. Assume we have a numeric vertex labeling such that each vertex is labeled to match the object it belongs to. The original BRT provided by Karras' algorithm is used to seperate vertices regardless of their label. Since our objective is to resolve between objects of different labels, we can proactively prune Karras' initial BRT, and subsequently the initial quadtree (see figure \ref{fig:pruning}) by allowing the generated quadtree leaves to contain multiple vertices as long as those vertices have the same label.

@@

Let the ``quadtree address'' refer to the unique ID of a quadtree cell $C$ found by concatenating the local addresses of its ancestors from Root to $C$, where the local address is a 2-bit (3-bit in 3D) Morton code. The address of the root cell is defined as the empty string. Figure \ref{fig:conflict-find-4} shows the address of each leaf cell in a quadtree.

>>>>

\paragraph{Parallel} Many recent works on fast quadtree construction using the GPU are limited to point sites \cite{bedorf2012sparse,karras2012maximizing,zhou2011data}.
%Kim and Liu's work~\shortcite{kim2014exact} is similar, computing the quadtree on the barycenters of triangles, giving an approximation of an object-resolving quadtree.
 Most quadtree approaches that support surfaces \cite{baert2013out,crassin2009gigavoxels,laine2011efficient,lefebvre2007compressed} are designed for efficient rendering, and actual construction of the quadtree is implemented on the CPU. Two works \cite{bastos2008gpu,park2010cuda} implement Adaptive Distance Fields in parallel on quadtrees but building the quadtree itself is done sequentially.  Yin \etal~\shortcite{yin2011fast} compute the octree entirely on the GPU using a bottom-up approach by initially subdividing into a complete quadtree, resulting in memory usage that is no better than using a uniform grid.
The most similar work to what we do here is Kim and Liu's method~\shortcite{kim2014exact}, which computes the quadtree on the barycenters of triangles, giving an approximation of our quadtree, but without fully resolving between objects.
We are unaware of any GPU quadtree construction methods that are fully adaptive and resolve the seperation between objects.

@@

\subsection{Build initial quadtree}
\label{sec:build-initial-quadtree}
Our first step is to build a quadtree on the given set of vertices. We use the Karras algorithm \cite{karras2012maximizing} which begins by placing the given vertices on a Z-Order curve by computing each vertex's cooresponding Morton code in parallel. Next, Karras sorts the converted points by using a parallel radix sorter, which has a linear execution time. Our implementation uses the efficient four-way parallel radix sorter described by Ha \etal \shortcite{ha2009fast}. Once the Morton codes are sorted, the Z-Order curve can be exploited to construct a binary radix tree in a parallel bottom up manner by identifying longest common Morton code prefixes between neighboring points. This resultant binary radix tree can be analyzed in parallel to identify the size and structure of the required vertex quadtree. The strength of this approach lies in the fact that overall performance scales linearly with the number of cores, regardless of the distribution of points. That is, even if a large number of vertices are clustered in a small area, requiring deep quadtree subdivision, only a constant number of parallel calls need be made.

@@

\subsection{Pruning the quadtree}
\label{sec:pruning}
During Karras' initial binary radix tree (BRT) construction, we can prune the BRT to simplify the resultant quadtree. This in turn simplifies the work complexity of conflict cell detection and reduces our overall memory footprint. Assume we have a numeric vertex labeling such that each vertex is labeled to match the object it belongs to. The original BRT provided by Karras' algorithm is used to generate a quadtree which separates vertices regardless of their label. Since our objective is to resolve between objects of different labels, we can proactively prune Karras' initial BRT, and subsequently the initial quadtree (see figure \ref{fig:pruning}) by allowing the generated quadtree leaves to contain multiple vertices as long as those vertices have the same label.

@@

Let the ``quadtree address'' refer to the unique ID of a quadtree cell $C$ found by concatenating the local addresses of its ancestors from the root quadtree cell to cell $C$, where the local address is a 2-bit (3-bit in 3D) Morton code. The address of the root cell is defined as the empty string. Figure \ref{fig:conflict-find-4} shows the address of each leaf cell in a quadtree.
